### 2025-10-04 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
2025NSF: typos and minor corrections

Physics is currently a patchwork of different theories with no common underlying foundation. As part of our larger project, Assumptions of Physics, we aim to develop an axiomatic basis for states and processes that can be equally applied to different theories, such as classical and quantum mechanics. Different theory are then recovered as specializations of the more general structure.

A core insight that is guiding this enterprise is that geometric structures in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two dimensional symplectic manifold. In this case, symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduce to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if the independent degrees of freedom are independently distributed. Therefore, if the distribution of all possible marginals are specified, the symplectic form can be recovered.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, mathematically $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi,\psi)}}{2},\frac{1 - \sqrt{p(\phi,\psi)}}{2} \right)$ where $I(x,y) = -x \log x -y \log y$ is the Shannon entropy. Therefore the entropy is an invertible function of $p(\psi,\phi)$. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the state space.

While the states spaces of classical (i.e. symplectic manifolds) and quantum mechanics (i.e. projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly convex entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year we have done significant preliminary work that shows that not only this approach is possible, but is extremely productive. The proposed work is to turn those preliminary insight in an axiomatic specification that is fully physically motivated. If successful, such an approach will provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories as well as the search for new ones;
	\item provide a generalized structure for information theory, possibly allowing future generalization of results from both classical and quantum information theory
	\item provide useful constrains to more precisely formulate physical theories in the infinite dimensional case.
\end{enumerate}

@@

We want to stress that finding the correct starting point is not just a mathematical issue, but also a physics problem. Let us explore this problem through an example. The space $M_1([0,1])$ of probability measures defined over the $[0,1]$ interval is an infinite dimensional vector space. As Aliprantis and Border note: ``Since there is more than one topology of interest on an infinite dimensional space, the choice of topology is a key modeling decision that can have economic as well as technical consequences.'' Unbeknowst to many physicists, this is true for physics as well. If we choose a weak topology, we are saying the convergence of expectation values of all continuous functions is enough to determine convergence of ensembles. While this may seem plausible at first glance, we can find a weakly convergence sequence for which the entropy does not converge. If the entropic structure of a physical theory is really equivalent to its geometric structure, not only the entropy should be continuous, but it is likely that the topology should be the one generated by the entropic structure. Ultimately, answering these questions means understanding both the physical and mathematical consequences of those decisions.

Our preliminary work convinced us that the notion of statistical ensemble (i.e. the collection of all outputs of a preparation procedure) seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the math, the structure of statistical ensembles of both classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this outcome.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that a physical theory must \emph{at least} posit a space of ensembles. Our preliminary work shows that there are three, possibly four, distinct set of requirements for an ensemble space $\Ens$, the set of ensemble given by a physical theory.

The first requirement is that of experimental verifiability of the theory. This has been thoroughly investigated in our previous work that established a precise correspondence between experimental verifiable statements (i.e. statement associated with an experimental test that terminates successfully in finite time if and only if the statement is true) and the open set of a ``natural'' topology, and between theoretical statements (i.e. statement associated with an experimental test without termination guarantees) and Borel sets. Real and integer numbers, for example, are recovered from an idealized model representing references of a reference system. The key insight is that the set theoretic operations represent a two-valued logic system (i.e. truth or falsehood of statements) while the topology keeps track of a three-valued logic system (i.e. interior, exterior and boundary represent respectively successful termination, unsuccessful termination and non-termination not terminate). This work gives us a solid foundation on which to build, and guarantees that an ensemble space $\Ens$ is, at least, a $\mathsf{T}_0$ second countable topological space.

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This is insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. While we haven't found such a structure already studied in the literature, and it would extremely useful to us to see what can be generalized from topological vector spaces to topological convex spaces, this is a fairly well defined mathematical question at this point, and we believe that our expertise in the mathematical foundations of physical theories is better spent in continuing finding other well defined starting points.

@@

The hardest conceptual problem we found is that the entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary work, for example, the entropy $S : \Ens \to \mathbb{R}$ has the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e. the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensemble, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be $ - \kappa \left(p \log p + \bar{p} \log \bar{p}\right)$, the Shannon entropy. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is valid with an equal, and can be used to calculate the entropy from that decomposition. That allows to recover the expression for the entropy in the three different cases.

Our preliminary work shows that an entropy satisfying those axioms forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that there exists an embedding $\iota : \Ens \hookrightarrow  V_{\ens}$ such that $\iota(p \ens[a] + \bar{p} \ens[b]) = p \iota(\ens[a]) + \bar{p} \iota(\ens[b])$ and $\iota(\ens) = 0$. If we pick two different internal points $\ens[a],\ens[b] \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ is $V_{\ens[b]}$. That is, the ensemble space embeds in an affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is finite. It is an open question whether the axioms are enough to prove that the embedding is continuous in a topological vector space, and whether the set is bounded.

We can also show that the entropy can be used to define geometric structures directly on the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{enumerate}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{enumerate}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical and quantum mechanics. The JSD is the square of a distance function in both classical and quantum information theory, though our current axioms do not guarantee $MS$ to be so. It is an open question whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls, and show that the topology must be Hausdorff. An interesting conjecture is whether an Hausdorff second countable topological convex space is metrizable, given that this is true for topological vector spaces. The current axioms do not guarantee the open balls to be convex. If we were to find sufficient physically motivated conditions that imply that the entropic open balls are convex and generate the topology, as we suspect it should be, the topology would be locally convex and therefore the entropic structure also gives us enough structure to define calculus.

>>>>

Physics is currently a patchwork of different theories with no common underlying foundation. As part of our larger project, Assumptions of Physics, we aim to develop an axiomatic basis for states and processes that can be equally applied to different theories, such as classical and quantum mechanics. Different theories are then recovered as specializations of the more general structure.

A core insight that is guiding this enterprise is that geometric structures in both classical and quantum mechanics are equivalent to entropic structures. In classical mechanics, a single degree of freedom is a two dimensional symplectic manifold. In this case, the symplectic form $\omega$ is a volume form and therefore specifying $\omega$ is equivalent to specifying the Liouville measure $\mu(U) = \int_U \omega$. In classical statistical mechanics, the entropy of a statistical ensemble is given by $S(\rho) = - \int \rho \log \rho d\mu$ where $\rho$ is the probability density associated to the ensemble. For a uniform distribution with support $U$, the entropy reduces to $S(\rho_U) = \log \mu(U)$. Since the logarithm is an invertible function, one can recover the measure from the entropy. That is, specifying the entropy associated to all uniform distributions allows us to reconstruct the Liouville measure, and therefore the symplectic form. In the general case, Darboux coordinates allow, at least locally, to factorize the space into multiple independent degrees of freedom, where the volume is the product of the corresponding areas. The corresponding entropic property is that the entropy for the joint distribution is the sum of the marginals if the independent degrees of freedom are independently distributed. Therefore, if the distribution of all possible marginals are specified, the symplectic form can be recovered.

In quantum mechanics, the geometry is given by the Born rule
\begin{equation}
	p(\phi|\psi) = \frac{\< \phi | \psi \>\< \psi | \phi \>}{\< \psi | \psi \>\< \phi | \phi \>}.
\end{equation}
Given two states $\psi$ and $\phi$, the ensemble given by their equal mixture, mathematically $\rho_{\psi\phi} = \frac{1}{2}|\psi\>\<\psi|+\frac{1}{2}|\phi\>\<\phi|$,  has von Neumann entropy $S(\rho_{\psi\phi}) = I\left(\frac{1 + \sqrt{p(\phi|\psi)}}{2},\frac{1 - \sqrt{p(\phi|\psi)}}{2} \right)$ where $I(x,y) = -x \log x -y \log y$ is the Shannon entropy. Therefore the entropy is an invertible function of $p(\phi|\psi)$. This means that, given the entropy of the equal mixtures of all pairs of states we can reconstruct the Born rule and, therefore, the state space.

While the state spaces of classical (i.e. symplectic manifolds) and quantum mechanics (i.e. projective Hilbert spaces) are rather different, the space of ensembles of both have a similar structure. They are both convex subsets of a topological vector space, equipped with a strictly concave entropy function. The core idea, then, is to use ensembles as a way to generalize the notion of physical states, and the entropy to characterize the geometry in a way that is independent of the particulars of each physical theory. In the past year we have done significant preliminary work that shows that not only this approach is possible, but is extremely productive. The proposed work is to turn those preliminary insights in an axiomatic specification that is fully physically motivated. If successful, such an approach will provide the following benefits:
\begin{enumerate}
	\item clarify exactly what physical premises guarantee the basic geometric structures of physical theories, allowing a more principled formulation of current physical theories as well as the search for new ones;
	\item provide a generalized structure for information theory, possibly allowing future generalization of results from both classical and quantum information theory
	\item provide useful constraints to more precisely formulate physical theories in the infinite dimensional case.
\end{enumerate}

@@

We want to stress that finding the correct starting point is not just a mathematical issue, but also a physics problem. Let us explore this problem through an example. The space $M_1([0,1])$ of probability measures defined over the $[0,1]$ interval is an infinite dimensional vector space. As Aliprantis and Border note: (add reference) ``Since there is more than one topology of interest on an infinite dimensional space, the choice of topology is a key modeling decision that can have economic as well as technical consequences.'' Unbeknowst to many physicists, this is true for physics as well. If we choose a weak topology, we are saying the convergence of expectation values of all continuous functions is enough to determine convergence of ensembles. While this may seem plausible at first glance, we can find a weakly convergent sequence for which the entropy does not converge. If the entropic structure of a physical theory is really equivalent to its geometric structure, not only the entropy should be continuous, but it is likely that the topology should be the one generated by the entropic structure. Ultimately, answering these questions means understanding both the physical and mathematical consequences of those decisions.

Our preliminary work convinced us that the notion of statistical ensemble (i.e. the collection of all outputs of a preparation procedure) seems to be particularly suitable for generalization for a number of both practical and conceptual considerations. In terms of the math, the structure of statistical ensembles of both classical and quantum mechanics can be made to look very similar. From the experimental side, statistical ensembles are what we produce and measure experimentally. On the conceptual side, the reproducibility requirement of scientific investigation means that physical laws are of the form ``whenever I prepare a system according to this procedure, I will find this range of outcomes.'' That is, physical laws are relationships between ensembles. Therefore, it makes sense to require that a physical theory must \emph{at least} posit a space of ensembles. Our preliminary work shows that there are three, possibly four, distinct set of requirements for an ensemble space $\Ens$, the set of ensemble given by a physical theory.

The first requirement is that of experimental verifiability of the theory. This has been thoroughly investigated in our previous work that established a precise correspondence between experimental verifiable statements (i.e. statement associated with an experimental test that terminates successfully in finite time if and only if the statement is true) and the open set of a ``natural'' topology, and between theoretical statements (i.e. statement associated with an experimental test without termination guarantees) and Borel sets. (add citation) Real and integer numbers, for example, are recovered from an idealized model representing references of a reference system. The key insight is that the set theoretic operations represent a two-valued logic system (i.e. truth or falsehood of statements) while the topology keeps track of a three-valued logic system (i.e. interior, exterior and boundary represent respectively successful termination, unsuccessful termination and non-termination). This work gives us a solid foundation on which to build, and guarantees that an ensemble space $\Ens$ is, at least, a $\mathsf{T}_0$ second countable topological space.

The second requirement is the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This is insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs). The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space. While we haven't found such a structure already studied in the literature, and it would be extremely useful to us to see what can be generalized from topological vector spaces to topological convex spaces, this is a fairly well defined mathematical question at this point, and we believe that our expertise in the mathematical foundations of physical theories is better spent in continuing finding other well defined starting points.

@@

The hardest conceptual problem we found is that the entropy is defined differently in each space. In a classical discrete space, the entropy is $-\sum p_i \log p_i$; in classical statistical mechanics is $- \int \rho \log \rho d\mu$ where $\mu$ is the Liouville measure; in quantum mechanics is $-\tr(\rho \log \rho)$. How can we give a single definition for these very different expressions? Note that the expressions are different precisely because they are in terms of the specific representation of ensembles in the specific theory. Our approach is to simply characterize the entropy by its properties with respect to the topology and the convex structure. In our preliminary work, for example, the entropy $S : \Ens \to \mathbb{R}$ has currently the following properties:
\begin{itemize}
	\item Continuity
	\item Strict concavity: $S(p\ens[a] + \bar{p} \ens[b]) \geq p S(\ens[a]) + \bar{p} S(\ens[b])$ with the equality holding if and only if $\ens[a] = \ens[b]$
	\item Upper variability bound: there exists a universal function $I(p_1, p_2)$ (i.e. the same for all ensemble spaces) such that $S(p\ens[a] + \bar{p} \ens[b]) \leq I(p, \bar{p}) + p S(\ens[a]) + \bar{p} S(\ens[b])$; if the equality holds, $\ens[a]$ and $\ens[b]$ are orthogonal, noted $\ens[a] \ortho \ens[b]$
	\item Mixtures preserve orthogonality: $\ens[a] \ortho \ens[b]$ and $\ens[a] \ortho \ens[c]$ if and only if $\ens[a] \ortho p \ens[b] + \bar{p} \ens[c]$ for any $p \in (0,1)$
\end{itemize}
These properties are justified by the physics. For example, if we are performing a statistical mixture between two ensembles, the average variability cannot decrease, which justifies the concavity. These relatively simple properties allow us, for example, to show that $I(p,\bar{p})$, the maximum entropy increase during mixing, must be, up to a multiplicative constant, the Shannon entropy $ - \left(p \log p + \bar{p} \log \bar{p}\right)$. The proof is general, meaning that if we are right that those axioms must hold for any physical theory, then the Shannon entropy is of truly general applicability. Moreover, in the case of both classical and quantum mechanics, states can be decomposed into orthogonal components, which means the upper variability is valid with an equal, and can be used to calculate the entropy from that decomposition. That allows to recover the expression for the entropy in the three different cases.

Our preliminary work shows that an entropy satisfying those axioms forces the convex space to be cancellative, meaning that it embeds into a vector space. That is, if we pick an internal point $\ens \in \Ens$, we can define a vector space $V_{\ens}$ such that there exists an embedding $\iota : \Ens \hookrightarrow  V_{\ens}$ such that $\iota(p \ens[a] + \bar{p} \ens[b]) = p \iota(\ens[a]) + \bar{p} \iota(\ens[b])$ and $\iota(\ens) = 0$. If we pick two different internal points $\ens[a],\ens[b] \in \Ens$, we can show that there is an invertible affine map between $V_{\ens[a]}$ and $V_{\ens[b]}$. That is, the ensemble space should be more properly thought as embedding in a affine space. We can also show that it is ``directionally bounded,'' meaning its intersection with any affine line is finite interval. It is an open question whether the current axioms are enough to prove that the embedding is continuous in a topological vector space, and whether the ensemble space is a bounded subset.

We can also show that the entropy can be used to define geometric structures directly on the ensemble space. For example, we can define the mixing entropy as $MS(\ens[a], \ens[b]) = S\left(\frac{1}{2}\ens[a] + \frac{1}{2} \ens[b]\right) - \left(\frac{1}{2} S(\ens[a]) + \frac{1}{2} S(\ens[b])\right)$. If we express the entropy in bits, the mixing entropy satisfies the following:
\begin{enumerate}
	\item non-negativity: $MS(\ens[a], \ens[b]) \geq 0$
	\item identity of indiscernibles: $MS(\ens[a], \ens[b]) = 0 \iff \ens[a]=\ens[b]$
	\item unit boundedness: $MS(\ens[a], \ens[b]) \leq 1$
	\item maximality of orthogonals: $MS(\ens[a], \ens[b]) = 1 \iff \ens[a] \ortho \ens[b]$
	\item symmetry: $MS(\ens[a], \ens[b]) = MS(\ens[b], \ens[a])$
\end{enumerate}
It is almost a distance function, as it does not satisfy the triangle inequality, but it recovers the Jensen-Shannon divergence (JSD) in both classical and quantum mechanics. While JSD is the square of a distance function in both classical and quantum information theory, our current axioms do not guarantee $MS$ to have this property in general. It is an open question whether enough physical constraints exist for that to be the case. The $MS$ can be used to create open balls, and show that the topology must be Hausdorff. An interesting conjecture is whether a Hausdorff second countable topological convex space is metrizable, given that this is true for topological vector spaces. The current axioms do not guarantee the open balls to be convex. If we were to find sufficient physically motivated conditions that imply that the entropic open balls are convex and generate the topology, as we suspect it should be, the topology would be locally convex and therefore the entropic structure also may give us enough structure to define calculus.
